RVD2: An ultra-sensitive variant detection model for low-depth heterogeneous next-generation sequencing data

### 2014-06-02 doc/bioinformatics2014/revision1/document.tex
revision1: Minor modification to the manuscript.

However, the primary statistical tools for calling variants from NGS data are optimized for homogeneous samples. Samtools/bcftools and GATK uses naive Bayesian decision rule to call variants~\citep{li2011statistical, depristo2011framework}. GATK involves more sophisticate pre- and post-processing steps wherein the genotype prior is fixed and constant across all loci and the likelihood of an allele at a locus is a function of the phred score~\citep{McKenna:2010bv}.

Recently, some have developed algorithms to call low-frequency or rare variants in heterogeneous samples.~\citet{yau2010statistical} developed a Bayesian framework which can model the normal DNA contamination and intra-tumor heterogeneity by parameterizing the normal genotype cell proportion at each SNP. VarScan2 combines algorithmic heuristics to call genotypes in the tumor and normal sample pileup data and then applies a Fisher's exact test on the read count data to detect a significant difference in the genotype calls~\citep{Koboldt:2012cg}. Strelka uses a hierarchical Bayesian approach to model the joint distribution of the allele frequency in the tumor and normal samples at each locus~\citep{Saunders:2012fh}. With the joint distribution available, one is able to identify locations with dissimilar allele frequencies. muTect uses a Bayesian posterior probability in its decision rule to evaluate the likelihood of a mutation~\citep{Cibulskis:2013ta}. RVD uses a hierarchical Bayesian model to capture the error structure of the data and call variants~\citep{Flaherty:2011ja, cushing2013rvd}.

@@

\begin{figure}[!bpth]
\centering
\vspace{-10pt}
\includegraphics[width=35mm]{pdf_figs/RVD2_model.pdf}
\caption{RVD2 Graphical Model.}
\vspace{-20pt}
\label{fig:graphical_model}
\end{figure}

@@

The joint distribution over the latent and observed variables for data at location $j$ in replicate $i$ given the parameters can be factorized as
\begin{multline}\label{eqn:jointpdf}
p \left( r_{ji}, \theta_{ji}, \mu_j | n_{ji}; \mu_0, M_0, M_j \right) \\ = p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right),
\end{multline}
where,
\begin{align}\nonumber
  p\left( \mu_j; \mu_0, M_0 \right)  &= \dfrac{\mu_j^{M_0\mu_0 -1} (1 - \mu_j)^{M_0 ( 1 - \mu_0) - 1}}{\textbf{B}(M_0 \mu_0,M_0 (1-\mu_0) )}, \nonumber \\
 %
  p\left( \theta_{ji} | \mu_j; M_j \right) &= \dfrac{ \theta_{ji}^{M_j\mu_j -1} (1 - \theta_{ji})^{M_j ( 1 - \mu_j) - 1}}{\textbf{B}(M_j \mu_j, M_j (1-\mu_j))}, \nonumber \\
 %
  p\left( r_{ji} | \theta_{ji}, n_{ji} \right) &= \frac{ \Gamma(n_{ji}+1) } { \Gamma(r_{ji}+1) \Gamma( n_{ji} - r_{ji} + 1 ) } \cdot \theta_{ji}^{r_{ji}} (1 - \theta_{ji})^{n_{ji} - r_{ji}}. \nonumber
\end{align}

%Integrating over the latent variables $\theta_{ji}$ and $\mu_j$ yields the marginal distribution of the data,
%\begin{multline}
%p \left( r_{ji} | n_{ji} ; \mu_0, M_0, M_j \right) = \int_{\mu_j} \int_{\theta_{ji}}  p \left( r_{ji} | \theta_{ji}, n_{ji} \right) \\ p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right) d\theta_{ji} d\mu_j.
%\end{multline}

>>>>

However, the primary statistical tools for calling variants from NGS data are optimized for homogeneous samples. Samtools and GATK uses naive Bayesian decision rule to call variants~\citep{li2011statistical, depristo2011framework}. GATK involves more sophisticate pre- and post-processing steps wherein the genotype prior is fixed and constant across all loci and the likelihood of an allele at a locus is a function of the phred score~\citep{McKenna:2010bv}.

Recently, some have developed algorithms to call low-frequency or rare variants in heterogeneous samples.~\citet{yau2010statistical} developed a Bayesian framework which can model the normal DNA contamination and intra-tumor heterogeneity by parameterizing the normal genotype cell proportion at each SNP. VarScan2 combines algorithmic heuristics to call genotypes in the tumor and normal sample pileup data and then applies a Fisher's exact test on the read count data to detect a significant difference in the genotype calls~\citep{Koboldt:2012cg}. Strelka uses a hierarchical Bayesian approach to model the joint distribution of the allele frequency in the tumor and normal samples at each locus~\citep{Saunders:2012fh}. With the joint distribution available, one is able to identify locations with dissimilar allele frequencies. muTect uses a Bayesian posterior probability in its decision rule to evaluate the likelihood of a mutation~\citep{Cibulskis:2013ta}. RVD uses a hierarchical Bayesian model to capture the error structure of the data and call variants~\citep{Flaherty:2011ja, cushing2013rvd}. That algorithm requires a very high read depth to estimate the sequencing error rate and call variants.

@@

\begin{figure}[!bpth]
\centering
\vspace{-10pt}
\includegraphics[width=33mm]{pdf_figs/RVD2_model.pdf}
\caption{RVD2 Graphical Model.}
\vspace{-20pt}
\label{fig:graphical_model}
\end{figure}

@@

The joint distribution over the latent and observed variables for data at location $j$ in replicate $i$ given the parameters can be factorized as
\begin{multline}\label{eqn:jointpdf}
p \left( r_{ji}, \theta_{ji}, \mu_j | n_{ji}; \mu_0, M_0, M_j \right) \\ = p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right),
\end{multline}

\begin{align}\nonumber
  p\left( r_{ji} | \theta_{ji}, n_{ji} \right) &= \frac{ \Gamma(n_{ji}+1) } { \Gamma(r_{ji}+1) \Gamma( n_{ji} - r_{ji} + 1 ) } \cdot \theta_{ji}^{r_{ji}} (1 - \theta_{ji})^{n_{ji} - r_{ji}}, \nonumber \\
 %
  p\left( \theta_{ji} | \mu_j; M_j \right) &= \dfrac{ \theta_{ji}^{M_j\mu_j -1} (1 - \theta_{ji})^{M_j ( 1 - \mu_j) - 1}}{\textbf{B}(M_j \mu_j, M_j (1-\mu_j))}, \nonumber \\
 %  
  p\left( \mu_j; \mu_0, M_0 \right)  &= \dfrac{\mu_j^{M_0\mu_0 -1} (1 - \mu_j)^{M_0 ( 1 - \mu_0) - 1}}{\textbf{B}(M_0 \mu_0,M_0 (1-\mu_0) )}, \nonumber 
\end{align}

where \textbf{B}() denotes the beta function.
%Integrating over the latent variables $\theta_{ji}$ and $\mu_j$ yields the marginal distribution of the data,
%\begin{multline}
%p \left( r_{ji} | n_{ji} ; \mu_0, M_0, M_j \right) = \int_{\mu_j} \int_{\theta_{ji}}  p \left( r_{ji} | \theta_{ji}, n_{ji} \right) \\ p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right) d\theta_{ji} d\mu_j.
%\end{multline}
